---
title: "Time Series"
subtitle: "Midterm project"
author: Sven Bergmann
date: 03/12/24
output: pdf_document
---

```{r, echo = F}
library(knitr)
opts_chunk$set(tidy.opts = list(width.cutoff = 60),
               tidy = TRUE)
```

# Part 1

## 1.

I am expecting a somewhat clear upward trend due to the increase in the population.
Also, there could be a seasonal trend, such that the retail sales are higher on special holidays.
Since the Brexit was later than 2007, I am not expecting a big sink somewhere in the graph.

```{r, warning = F, echo = F}
library(TSA)
```
```{r}
data(retail)
plot(retail, xlab = expression("Time"), ylab = expression("Retail sales"), main = expression("Time Series Plot of Retail sales"))
points(retail, pch = as.vector(season(retail)))
```

Seeing the actual graph, I can confirm the upward trend and also a very strong seasonal behaviour.
I clearly expected more holidays to be that present as christmas.
There surely is a really strong seasonal behaviour due to christmas.
This means that the sales are going up starting November each year and reach the peak in December before dropping again.

## 2.

I would try a seasonal ARIMA model, given the upward trend and the potential seasonality.

```{r}
model <- forecast::auto.arima(retail, seasonal = TRUE)
summary(model)
```
