---
title: "Problem Solving Set 7"
author: Sven Bergmann
date: March 11, 2024
output: pdf_document
---

```{r, echo = F}
library(knitr)
opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE, fig.width = 12, fig.height = 8)
```
```{r}
library(TSA)
library(MASS)
```

# Problem 2

```{r}
wn <- c(0.63, -1.25, 1.80, 1.51, 1.56, 0.62, 0.64, -0.98)
Y <- wn[1]
for (i in 2:length(wn)) {
  Y <- c(Y, 3 * Y[i - 1] + wn[i])
  print(Y)
}
```

# Problem 5

```{r}
data(airpass)
```

## a)

```{r}
plot(airpass, xlab = "Time", ylab = "Airline Passengers", main = "Time Series Plot of Passengers")
points(y = airpass, x = time(airpass), pch = as.vector(season(airpass)))
```

- seasonality (peaks in warmer months)
- upward linear trend
- increasing spread over time
-> not stationary

## b)

```{r}
logs <- as.ts(log(airpass))
plot(logs, xlab = "Time", ylab = "Log Passengers", main = "Plot of Log Passengers")
points(y = logs, x = time(logs), pch = as.vector(season(logs)))
```

- positive trend
- seasonality
- spread fairly even now

## c)

```{r}
diffs <- diff(airpass, 1)
airpass.2 <- ts(airpass[-1], start = c(1960, 2), end = c(1971, 12), frequency = 12)
frac.rel <- ts(diffs / airpass.2, start = c(1960, 2), end = c(1971, 12), frequency = 12)
```

```{r}
plot(frac.rel, xlab = "Time", ylab = "Fractional Relative Difference", main = "Time Series Plot--Fractional Rel Diff")
points(y = frac.rel, x = time(frac.rel), pch = as.vector(season(frac.rel)))
```

$\frac{Y_t - Y_{t-1}}{Y_{t-1}}$

- seasonality
- "random" scatter about 0
- fairly even spread

```{r}
plot(diff(logs), xlab = "Time", ylab = "First difference--Logs", main = "First Difference--Logs")
points(y = diff(logs), x = time(diff(logs)), pch = as.vector(season(diff(logs))))
```

$\bigtriangledown \log(Y_t)$

- nearly indistinguishable
- from fractional relative difference

## Fit seasonal means model to transformed series

```{r}
dif <- diff(logs)
months <- season(diff(logs))
model <- lm(dif ~ months)
```

### Modeling $Y_t$

1. Use $\log(Y_t)$, remove linear trend
- fit seasonal means or use cosine trend
2. Use frac.rel.diff or $\bigtriangledown\log(Y_t)$
- fit seasonal means or use cosine trend

# Problem 6

```{r}
data(JJ)
```

```{r}
plot(JJ, xlab = "Time", ylab = "Earnings Per Share", main = "Time Series Plot of Quarterly Earnings Per Share")
points(y = JJ, x = time(JJ), pch = as.vector(season(JJ)))

```

- exponential or quadratic positive trend
- seasonality (peak in Q3, dips in Q4)
- variance increases over time:
-> not stationary

```{r, warning = F}
lam <- seq(-0.5, 0.5, 0.1)
m <- BoxCox.ar(JJ, lambda = lam)$loglike
```

### Box-Cox transformation

$y^* = \frac{y^\lambda - 1}{\lambda}$

```{r, warning = F}
lambda <- lam[which(m == max(m))]
transform <- (JJ^lambda - 1) / lambda
```

For $\hat{\lambda} = 0.2$:
$y^* = \frac{y^{0.2}-1}{0.2}$

```{r}
plot(transform, xlab = "Time", ylab = "Transformed Earnings", main = "Time Series Plot--Transformed Earnings")
points(y = transform, x = time(transform), pch = as.vector(season(transform)))
```

- trend is now linear
- variance is a little more stable, still a bit uneven
- still seasonality

```{r}
diffs <- diff(transform, 1)
plot(diffs, xlab = "Time", ylab = "First Difference--Transformed Earnings", main = "First Difference--Transformed Earnings")
points(y = diffs, x = time(diffs), pch = as.vector(season(diffs)))
```

### Possible fixes

- take difference to remove the linear trend
- cosine trend or seasonal means to address seasonality

### What we actually see
- difference took out the linear trend

```{r}
diffs2 <- diffs + 1 # get rid of negative values
logs2 <- log(diffs2)
months <- season(logs2)
plot(logs2, xlab = "Time", ylab = "Transformed Earnings", main = "Time Series Plot--Transformed Earnings")
points(y = logs2, x = time(logs2), pch = as.vector(season(logs2)))
```

- log transform fared no better
- still have the bowtie pattern
- still have the seasonality

```{r}
model2 <- lm(logs2 ~ months)
res <- ts(rstandard(model2), start = c(1960, 2), end = c(1980, 3), frequency = 4)
plot(res, xlab = "Time", ylab = "Residual", main = "Time Series Plot of Residuals")
points(y = res, x = time(res), pch = as.vector(season(res)))
```

- fitting seasonal means fixes seasonality
- variance problem remains

```{r}
plot(res, xaxt = "n", xlab = "Time", ylab = "Residual", main = "Time Series Plot of Residuals")
points(y = res, x = time(res), pch = as.vector(season(res)))
```

```{r}
par(mfrow = c(1, 2))
qqnorm(res, main = "Normal Probability Plot vs \n Residuals Fit")
abline(a = 0, b = 1, col = "red")
hist(res, xlab = "Residual", main = "Histogram of Residuals")
```
