---
title: "Problem Solving Set 1"
author: Sven Bergmann
date: January 24, 2024
output: pdf_document
---

```{r, echo = F}
library(knitr)
opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE, fig.width = 12, fig.height = 8)
```
```{r}
set.seed(1401)
```
# Part 1
```{r}
n <- 120
```
```{r}
series1 <- as.ts(rchisq(n, 2, ncp = 0))
series2 <- as.ts(rchisq(n, 2, ncp = 0))
series3 <- as.ts(rchisq(n, 2, ncp = 0))
series4 <- as.ts(rchisq(n, 2, ncp = 0))
series5 <- as.ts(rchisq(n, 2, ncp = 0))
```
```{r}
#pdf.options()
#pdf("PS1TSPlotsPt1.pdf")
par(mfrow = c(3, 2))
plot(series1, xlab = expression("Time"), ylab = expression(paste("Y"[1][t])), main = expression(paste("Time Series Plot of Y"[1[t]])))
plot(series2, xlab = expression("Time"), ylab = expression(paste("Y"[2][t])), main = expression(paste("Time Series Plot of Y"[2[t]])))
plot(series3, xlab = expression("Time"), ylab = expression(paste("Y"[3][t])), main = expression(paste("Time Series Plot of Y"[3[t]])))
plot(series4, xlab = expression("Time"), ylab = expression(paste("Y"[4][t])), main = expression(paste("Time Series Plot of Y"[4[t]])))
plot(series5, xlab = expression("Time"), ylab = expression(paste("Y"[5][t])), main = expression(paste("Time Series Plot of Y"[5[t]])))
#dev.off()
```

### $Y_1$
- spikes between 50 and 70
- centered around 2
- no patterns / random scatter

### $Y_2$
- more variation than $Y_1$
- still centered at 2
- no patterns

### $Y_3$
- similar to $Y_1$
- spike at 70

### $Y_4$
- centered at 2
- spike near 60
- random scatter

### $Y_5$
- similar to $Y_1$, $Y_3$
- spike at / around 90
```{r}
#pdf.options()
#pdf("PS1NPPlotsPt1.pdf")
par(mfrow = c(3, 2))
qqnorm(series1, main = expression(paste("Normal Probability Plot of Y"[1[t]])))
qqnorm(series2, main = expression(paste("Normal Probability Plot of Y"[2[t]])))
qqnorm(series3, main = expression(paste("Normal Probability Plot of Y"[3[t]])))
qqnorm(series4, main = expression(paste("Normal Probability Plot of Y"[4[t]])))
qqnorm(series5, main = expression(paste("Normal Probability Plot of Y"[5[t]])))
#dev.off()
```
- not linear,
- heavy right tail -> skewed right
```{r}
#pdf.options()
#pdf("PS1HistsPt1.pdf")
par(mfrow = c(3, 2))
hist(series1, xlab = expression(paste("Y"[1[t]])), main = expression(paste("Histogram of Y"[1[t]])))
hist(series2, xlab = expression(paste("Y"[2[t]])), main = expression(paste("Histogram of Y"[2[t]])))
hist(series3, xlab = expression(paste("Y"[3[t]])), main = expression(paste("Histogram of Y"[3[t]])))
hist(series4, xlab = expression(paste("Y"[4[t]])), main = expression(paste("Histogram of Y"[4[t]])))
hist(series5, xlab = expression(paste("Y"[5[t]])), main = expression(paste("Histogram of Y"[5[t]])))
#dev.off()
```
# Part 2

## Expectations

- properties of a t-dist.
- look somewhat normal
- unimodal
- symmetric
- centered about 0
- random scatter / no patterns - property of random sample
```{r}
n <- 100
series1 <- as.ts(rt(n, 5, ncp = 0))
series2 <- as.ts(rt(n, 5, ncp = 0))
series3 <- as.ts(rt(n, 5, ncp = 0))
series4 <- as.ts(rt(n, 5, ncp = 0))
series5 <- as.ts(rt(n, 5, ncp = 0))
```
```{r}
#pdf.options()
#pdf("PS1TSPlotsPt2.pdf")
par(mfrow = c(3, 2))
plot(series1, xlab = expression("Time"), ylab = expression(paste("Y"[1][t])), main = expression(paste("Time Series Plot of Y"[1[t]])))
plot(series2, xlab = expression("Time"), ylab = expression(paste("Y"[2][t])), main = expression(paste("Time Series Plot of Y"[2[t]])))
plot(series3, xlab = expression("Time"), ylab = expression(paste("Y"[3][t])), main = expression(paste("Time Series Plot of Y"[3[t]])))
plot(series4, xlab = expression("Time"), ylab = expression(paste("Y"[4][t])), main = expression(paste("Time Series Plot of Y"[4[t]])))
plot(series5, xlab = expression("Time"), ylab = expression(paste("Y"[5][t])), main = expression(paste("Time Series Plot of Y"[5[t]])))
#dev.off()
```

### $Y_1$
- centered at 0
- spike at 50
- random scatter / no patterns

### $Y_2$
- signs of uneven spread
- suggests nonstationary series
- centered at 0
- no patterns

### $Y_3$
- centered at 0
- random scatter
- fairly even spread

### $Y_4$
- centered at 0
- possible periodic pattern
- fairly even spread

### $Y_5$
- centered at 0
- random scatter
- fairly even spread
```{r}
#pdf.options()
#pdf("PS1NPPlotsPt2.pdf")
par(mfrow = c(3, 2))
qqnorm(series1, main = expression(paste("Normal Probability Plot of Y"[1[t]])))
qqnorm(series2, main = expression(paste("Normal Probability Plot of Y"[2[t]])))
qqnorm(series3, main = expression(paste("Normal Probability Plot of Y"[3[t]])))
qqnorm(series4, main = expression(paste("Normal Probability Plot of Y"[4[t]])))
qqnorm(series5, main = expression(paste("Normal Probability Plot of Y"[5[t]])))
#dev.off()
```

### $Y_1$
- heavy right tail
- left side - tail resembles that of a normal distribution

### $Y_2$
- two heavy tails
- middle looks really normal

### $Y_3$
- heavy tails
- middle looks okay
- lighter than tails of $Y_2$

### $Y_4$
- both tails are heavy
- middle is linear

### $Y_5$
- similar to $Y_4$
```{r}
#pdf.options()
#pdf("PS1HistsPt2.pdf")
par(mfrow = c(3, 2))
hist(series1, xlab = expression(paste("Y"[1[t]])), main = expression(paste("Histogram of Y"[1[t]])))
hist(series2, xlab = expression(paste("Y"[2[t]])), main = expression(paste("Histogram of Y"[2[t]])))
hist(series3, xlab = expression(paste("Y"[3[t]])), main = expression(paste("Histogram of Y"[3[t]])))
hist(series4, xlab = expression(paste("Y"[4[t]])), main = expression(paste("Histogram of Y"[4[t]])))
hist(series5, xlab = expression(paste("Y"[5[t]])), main = expression(paste("Histogram of Y"[5[t]])))
#dev.off()
```
