---
title: "Problem Solving Set 1"
author: Sven Bergmann
date: January 24, 2024
output: pdf_document
---

```{r, echo = F}
library(knitr)
opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE, fig.width = 12, fig.height = 8)
```
```{r}
set.seed(1401)
```
# Part 1
```{r}
n <- 120
```
```{r}
series_list <- replicate(5, as.ts(rchisq(n, 2, ncp = 0)))
```
```{r}
par(mfrow = c(3, 2))
for (i in 1:5) {
  plot(series_list[, i], type = 'l', xlab = expression("Time"), ylab = expression(paste("Y"[i][t])), main = expression(paste("Time Series Plot of Y"[1[i]])))
}
```

### $Y_1$
- spikes between 50 and 70
- centered around 2
- no patterns / random scatter

### $Y_2$
- more variation than $Y_1$
- still centered at 2
- no patterns

### $Y_3$
- similar to $Y_1$
- spike at 70

### $Y_4$
- centered at 2
- spike near 60
- random scatter

### $Y_5$
- similar to $Y_1$, $Y_3$
- spike at / around 90
```{r}
par(mfrow = c(3, 2))
for (i in 1:5) {
  qqnorm(series_list[, i], main = expression(paste("Normal Probability Plot of Y"[i[t]])))
  abline(a = 0, b = 1, col = 'red')
}
```
- not linear,
- heavy right tail -> skewed right
```{r}
par(mfrow = c(3, 2))
for (i in 1:5) {
  hist(series_list[, i], xlab = expression(paste("Y"[i[t]])), main = expression(paste("Histogram of Y"[i[t]])))
}
```
# Part 2

## Expectations

- properties of a t-dist.
- look somewhat normal
- unimodal
- symmetric
- centered about 0
- random scatter / no patterns - property of random sample
```{r}
n <- 100
```
```{r}
series_list <- replicate(5, as.ts(rt(n, 5, ncp = 0)))
```
```{r}
par(mfrow = c(3, 2))
for (i in 1:5) {
  plot(series_list[, i], type = 'l', xlab = expression("Time"), ylab = expression(paste("Y"[i][t])), main = expression(paste("Time Series Plot of Y"[1[i]])))
}
```

### $Y_1$
- centered at 0
- spike at 50
- random scatter / no patterns

### $Y_2$
- signs of uneven spread
- suggests nonstationary series
- centered at 0
- no patterns

### $Y_3$
- centered at 0
- random scatter
- fairly even spread

### $Y_4$
- centered at 0
- possible periodic pattern
- fairly even spread

### $Y_5$
- centered at 0
- random scatter
- fairly even spread
```{r}
par(mfrow = c(3, 2))
for (i in 1:5) {
  qqnorm(series_list[, i], main = expression(paste("Normal Probability Plot of Y"[i[t]])))
  abline(a = 0, b = 1, col = 'red')
}
```

### $Y_1$
- heavy right tail
- left side - tail resembles that of a normal distribution

### $Y_2$
- two heavy tails
- middle looks really normal

### $Y_3$
- heavy tails
- middle looks okay
- lighter than tails of $Y_2$

### $Y_4$
- both tails are heavy
- middle is linear

### $Y_5$
- similar to $Y_4$
```{r}
par(mfrow = c(3, 2))
for (i in 1:5) {
  hist(series_list[, i], xlab = expression(paste("Y"[i[t]])), main = expression(paste("Histogram of Y"[i[t]])))
}
```
