---
title: "Problem Solving Set 8"
author: Sven Bergmann
date: March 25, 2024
output: pdf_document
---

```{r, echo = F}
library(knitr)
opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE, fig.width = 12, fig.height = 8)
```
```{r, warning = F, message = F}
library(TSA)
library(tseries)
```
```{r}
set.seed(1172)
```

# Problem 7

```{r}
phi <- c(0.7, -0.4)
ar2.series <- arima.sim(n = 72, list(ar = phi, sd = 1))
```

## Part a

```{r}
ar2acf <- ARMAacf(ar = c(0.7, -0.4), lag.max = 10, pacf = FALSE)
plot(ar2acf, xlab = "Lag", ylab = "Theoretical ACF", main = "Theoretical ACF AR(2)", type = "h") # type = 'h' gives the vertical lines
abline(h = 0)
```

- dies out beyond lag 5

## Part b

```{r}
acf.samp <- acf(ar2.series, main = "AR(2) Sample Autocorrelation", lag.max = 40)
acf.samp$acf
```

- "cuts off" after lag 1
- dies out way slower than the theoretical autocorrelation
- due to short series / small sample size
- meaning if we take a longer series, e.g. 7200 it dies out way faster

## Part c

```{r}
rho12 <- ar2acf[2:3]
rho1 <- phi[1] / (1 - phi[2])
rho2 <- (rho12[2] - rho12[1]^2) / (1 - rho12[1]^2)
rho1
rho2
```
```{r}
ar2pacf <- ARMAacf(ar = c(0.7, -0.4), pacf = TRUE)
plot(ar2pacf, xlab = "Lag", ylab = "Theoretical PACF", main = "Theoretical PACF AR(2)", type = "h")
abline(h=0)
```

- PACF is denoted in the book also as $\phi$ but with to subscripts
- $\phi_{1,1} = \frac{\phi_1}{1-\phi_2}$
- $\phi_{2,2} = \frac{\rho_2 - \rho_1^2}{1 - \rho_1^2}$
- cuts off after lag 2 in theory

## Part d

```{r}
samp.pacf <- pacf(ar2.series, main = "AR(2) Sample Partial Autocorrelation")
```

- does not cut off after lag 2
- significant autocorrelation at lag 6
- does not match theoretical PACF well

```{r}
eacf(ar2.series)
```

# Problem 8

```{r}
theta <- c(-0.7, 0.4)
ma2 <- arima.sim(n = 36, list(ma = theta, sd = 1))
ma2acf <- ARMAacf(ma = theta, lag.max = 4, pacf = FALSE)
ma2acf
```
```{r}
acfs <- acf(ma2, main = "ACF for MA(2)")
acfs$acf
```

- matches theoretical ACF really well.

```{r}
acfma2 <- acfs$acf
```
```{r}
pacfMA2 <- ARMAacf(ma = theta, lag.max = 10, pacf = TRUE)
pacfMA2
plot(pacfMA2, xlab = "Lag", ylab = "Theoretical PACF", main = "Theoretical PACF MA(2)", type = "h")
abline(h = 0)
```

- dies out

```{r}
pacf.samp <- pacf(ma2, main = "Sample MA(2) PACF")
pacfs <- pacf.samp$acf
pacfs
```

- match up really well with the theoretical PACF

```{r}
#eacf(ma2)
```

### Specification

- fit MA(1), AR(1), ARMA(1,1)
- see what works best

# Problem 9

## Part a

```{r}
theor.ac <- ARMAacf(ar = c(0.7), lag.max = 5, pacf = FALSE)
theor.pac <- ARMAacf(ar = c(0.7), lag.max = 5, pacf = TRUE)
set.seed(2)
ar.series <- arima.sim(n = 48, list(ar = 0.7, sd = 1))
```

## Part b

```{r}
acs <- acf(ar.series, plot = F)$acf[c(1, 5)]
acs
```

## Part c

```{r}
ar.series2 <- arima.sim(n = 48, list(ar = 0.7, sd = 1))
acs2 <- acf(ar.series2, plot = F)$acf[c(1, 5)]
acs2
```

## Part d

```{r}
m <- 1000
ac1 <- NULL
ac5 <- NULL
for (i in 1:m) {
  ar.s <- arima.sim(n = 48, list(ar = 0.7, sd = 1))
  ac1 <- c(ac1, acf(ar.s, plot = F)$acf[1])
  ac5 <- c(ac5, acf(ar.s, plot = F)$acf[5])

}
par(mfrow = c(1, 1))
hist(ac1, xlab = expression(r[1]), main = expression(paste("Histogram of ", r[1])))
hist(ac5, xlab = expression(r[5]), main = expression(paste("Histogram of ", r[5])))
```
```{r}
var(ac1)
```
```{r}
var(ac5)
```
